---
title: "Reproducible Research - Course Project 1"
author: "S. Grath"
date: "November 24, 2017"
output: html_document
---

##Activity Monitoring Data
###Loading and preprocessing the data

```{r}
#Load activity data
myActivityData <- read.table("activity.csv", sep = ",", header = TRUE)
#Some controls
summary(myActivityData)
str(myActivityData)
names(myActivityData)
```

###What is mean total number of steps taken per day?

For this part of the assignment, I ignored the missing values in the dataset.

```{r}
#Calculate the mean - ignore na values
meanStepsDay <- tapply(myActivityData$steps, 
                       myActivityData$date, 
                       sum, 
                       na.rm = TRUE)
#Convert into data frame
meanStepsDay <- as.data.frame(as.matrix(tapply(myActivityData$steps, 
                                               myActivityData$date, 
                                               sum, 
                                               na.rm = TRUE)))
#Change column names
colnames(meanStepsDay) <- c("totalStepsDay")
#Print data frame
meanStepsDay

#Plot histogram
library(ggplot2)
ggplot(meanStepsDay, aes(x=totalStepsDay)) + 
    geom_histogram(binwidth = 1000, 
                   fill = "green", 
                   colour = "red", 
                   size=0.2) + 
    labs(title="Histogram for total number of steps per day") +
    labs(x="steps per day", y="count") + 
    theme(
        plot.title = element_text(color="black", 
                                  size=14, 
                                  face="bold", 
                                  hjust = 0.5)
    )
ggsave("1.Histogram.pdf")
#Calculate mean and median
meanSteps <- as.integer(mean(meanStepsDay$totalStepsDay))
medianSteps <- as.integer(median(meanStepsDay$totalStepsDay))
```

The average number of steps per day is `r meanSteps`. The median number of steps per day is `r medianSteps`. Note that I gave the mean and median numbers as integers. In my opinion, only full steps make sense here.

###What is the average daily activity pattern?

First, I made a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged over all days (y-axis). Second, I calculated the 5-minute interval, that, on average across all days in the dataset, contains the maximum number of steps.

```{r}
intervalSteps <- aggregate(myActivityData$steps, 
                           list(myActivityData$interval), 
                           mean, 
                           na.rm = TRUE)
colnames(intervalSteps) <- c("interval", "averageSteps")
ggplot(intervalSteps, aes(interval, averageSteps)) + 
    geom_line() +
    labs(title="Time series plot with average number of steps per interval") +
    labs(y="average steps", x="interval") + 
    theme(
        plot.title = element_text(color="black", 
                                  size=14, 
                                  face="bold", 
                                  hjust = 0.5)
    )
ggsave("2.TimeSeriesPlot.pdf")
mostSteps <- intervalSteps[which.max(intervalSteps$averageSteps), ]
myInterval <- mostSteps$interval
mySteps <- as.integer(mostSteps$averageSteps)
```

The **interval with the maximum number of steps** on average is: `r myInterval` <br>
This interval has on average `r mySteps` steps. Again, the step value is given as integer.

###Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.<br>
First, I calculated and reported the total number of missing values in the dataset (i.e. the total number of rows with NAs). Second, I defined a strategy for filling in all missing values in the dataset. I used the mean for any given 5-minute interval. Third, I created a new dataset that is equal to the original dataset but with the missing data filled in. Last, I draw a histogram of the total number of steps taken each day and calculated and reported the mean and median total number of steps taken per day.

```{r}
missingValues<- is.na(myActivityData$steps)
missingValuesTotal <- sum(as.numeric(missingValues))
```

The **total number of missing values** (rows with NA) is:** `r missingValuesTotal`

```{r}
for (i in 1:length(myActivityData$steps)){
    if (is.na(myActivityData$steps[i] == TRUE)){        
        myActivityData$steps[i] <- 
            intervalSteps$averageSteps[match(myActivityData$interval[i], 
                                             intervalSteps$interval)]  
    } 
}

meanStepsDayModified <- as.data.frame(as.matrix(tapply(myActivityData$steps, 
                                                       myActivityData$date, 
                                                       sum)))
colnames(meanStepsDayModified) <- c("totalStepsDayModified")

ggplot(meanStepsDayModified, aes(x=totalStepsDayModified)) + 
    geom_histogram(binwidth = 1000, 
                   fill = "green", 
                   colour = "red", 
                   size=0.2) + 
    labs(title="Histogram for total number of steps per day (data modified)") +
    labs(x="steps per day (modified)", y="count") + 
    theme(
        plot.title = element_text(color="black", 
                                  size=14, 
                                  face="bold", 
                                  hjust = 0.5)
    ) 
ggsave("3.HistogramModifiedData.pdf")
#Calculate mean and median
meanStepsModified <- as.integer(mean(meanStepsDayModified$totalStepsDayModified))
medianStepsModified <- as.integer(median(meanStepsDayModified$totalStepsDayModified))
```

The **mean number of steps** in the modified dataset is: `r meanStepsModified`.
The **median number of steps** in the modified dataset is: `r medianStepsModified`.<br>

To compare the original with the modified dataset, I draw histograms of the total number of steps per day for both datasets next to each other. You can see, that in the modified dataset the number of times with 0 steps per day decreases, while the numbers of times with average steps per day (e.g. 10 000 steps) increase.

```{r}
#Direct comparison of original data set with modified data set
combination <- merge(meanStepsDay, meanStepsDayModified)
library(reshape2)
combination <- melt(combination)
colnames(combination) <- c("type", "totalStepsDay")

labels <- c(totalStepsDay = "total steps per day", 
            totalStepsDayModified = "total steps per day (modified)")
ggplot(combination, aes(x=totalStepsDay)) + 
    geom_histogram(binwidth = 1000, 
                   fill = "orange", 
                   colour = "grey60", 
                   size=0.2) + 
    facet_grid(.~ type, 
               labeller = labeller(type = labels)) +
    labs(title="Total number of steps per day for both data sets") +
    labs(x="steps per day", y="count") + 
    theme(
        plot.title = element_text(color="black", 
                                  size=14, 
                                  face="bold", 
                                  hjust = 0.5)
    ) +
    theme(axis.title = element_text(face = "bold", 
                                    size = rel(1.0)),
          strip.text = element_text(face = "bold", 
                                    size = rel(1.0)),
          strip.background = element_rect(fill="lightblue", 
                                          colour = "black", 
                                          size = 1))
ggsave("4.HistogramsComparison.pdf")
```

###Are there differences in activity patterns between weekdays and weekends?

For this part of the assignment, I used the weekdays() function and the dataset with the filled-in missing values.<br>

```{r}
#Add the day-of-week column
myActivityData$day <- weekdays(as.Date(myActivityData$date), abbreviate=FALSE)

#Convert day-of-week column vales to Weekday or Weekend
#CAUTION! German system of date format - Samstag = Saturday, Sonntag = Sunday
#Script would work for both English and German day names
for (i in 1:length(myActivityData$day)){
    if (myActivityData$day[i] == "Samstag" 
        || myActivityData$day[i] == "Sonntag"
        || myActivityData$day[i] == "Saturday" 
        || myActivityData$day[i] == "Sunday"){
        myActivityData$day[i] <- "Weekend"
    } else {
        myActivityData$day[i] <- "Weekday"
    }
}

#Subset the data based on 'day' (Weekend or Weekday)
activityDataWeekend <- subset(myActivityData, day == "Weekend")
activityDataWeekday <- subset(myActivityData, day == "Weekday")

#Calculate the average interval steps for weekend and weekdays separetely
activityIntervalStepsWeekend <- aggregate(activityDataWeekend$steps, 
                                          list(activityDataWeekend$interval), 
                                          mean, 
                                          na.rm = TRUE)
activityIntervalStepsWeekday <- aggregate(activityDataWeekday$steps, 
                                          list(activityDataWeekday$interval), 
                                          mean)
#Change colnames
colnames(activityIntervalStepsWeekend) <- c("interval", "weekend")
colnames(activityIntervalStepsWeekday) <- c("interval", "weekday")

#Combine in one data frame
DF <- merge(activityIntervalStepsWeekday, activityIntervalStepsWeekend)
DF <- melt(DF, id = "interval")

#Plot
colnames(DF) <- c("interval", "type", "numberSteps")
ggplot(DF, aes(interval,  numberSteps)) + 
    geom_line(colour="#000099") + 
    facet_wrap(~ type, nrow = 2) +
    labs(x="Interval", y="Number of Steps") + 
    theme(axis.title = element_text(face = "bold", size = rel(1.0)),
          strip.text = element_text(face = "bold", size = rel(1.0)),
          strip.background = element_rect(fill="lightblue", 
                                          colour = "black", 
                                          size = 1))
ggsave("5.ComparisonWeekendWeekday.pdf")
```

```{r}
sessionInfo()
Sys.time()
```